%\pagewidth 210mm
%\pageheight 297mm
\documentclass[a4paper,10pt]{article}

\usepackage[pdfusetitle]{hyperref}

\usepackage[a4paper,margin=1in]{geometry}
\usepackage[bf,tiny]{titlesec}
\usepackage{fancyhdr}

%\usepackage[indent=0pt,skip=10pt]{parskip}

\usepackage{amsmath}
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\usepackage{amssymb}
\let\mathbbalt\mathbb

\usepackage{fontspec}
\usepackage{unicode-math}
\let\mathbb\mathbbalt

\newcommand{\naturals}{\mathbb{N}}
\newcommand{\reals}{\mathbb{R}}
\newcommand{\rationals}{\mathbb{Q}}
\newcommand{\integers}{\mathbb{Z}}

\newcounter{problemno}
\setcounter{problemno}{1}
\newcommand{\problem}{\section*{\textbf{Problem \theproblemno}}\refstepcounter{problemno}}
\newcounter{interludeno}
\setcounter{interludeno}{1}
\newcommand{\interlude}[1]{\par\noindent\\\fbox{\begin{minipage}{\linewidth}\textbf{Interlude \theinterludeno.\enspace}#1\end{minipage}}\\\refstepcounter{interludeno}}

\newcounter{questionno}
\setcounter{questionno}{1}
\newcommand{\question}[1]{\par\noindent{\thequestionno.\enspace}#1\refstepcounter{questionno}\\}

\newcommand{\thetitle}{Term 1 Week 10 Handout}

\renewcommand\thesection{\S{\arabic{section}}.}
\renewcommand\thesubsection{\S{\arabic{section}.\arabic{subsection}}}

\title{MEG 2023 Term 1 Week 10 Handout}
\author{Nathan Wong\and Tom Yan}
\date{2023}
\begin{document}
\noindent Melbourne High School\\
Maths Extension Group 2023\\
\textbf{\thetitle}\\
\section{Problems relating to combinatorial games}
\begin{enumerate}
\item A pile of $a$ stones is given. Two players play the following game: each of them in turns takes 1,2 or 3 stones. The player who takes the last stone wins. Determine, in terms of $a$, who has a winning strategy. 
\item  In the game $Double Chess$, the rules of chess are changed so that White and Black alternately make two legal moves at a time. Show that Black doesn’t have a winning strategy 
\item Two players start with the number 1 and take turns to multiply it by an integer from 2
to 9. The winner is the first player to obtain a number greater than or equal to 1000. Which player has a winning strategy? 
\item All cells of a $4 \times 4$ table are painted white. In turn two players paint a white cell black. If a player makes a move, after which there is a black square $2 \times 2$, the player loses. Who has a winning strategy?
\item Two players in turn write $S$ or $O$ in an empty cell of a $1 \times 2000$ table. The first player who produces three consecutive boxes that spell SOS wins. If all boxes are filled without producing SOS then the game is a draw. Prove that the second player has a winning strategy.
\end{enumerate}
\twocolumn
\input{congr.tex}
\end{document}