\documentclass[a4paper,12pt]{exam}
\usepackage{geometry}

\usepackage{fontspec}
\usepackage{unicode-math}
\setmainfont{TeX Gyre Termes}
\setmathfont{TeX Gyre Termes Math}

\begin{document}
\begin{questions}
  \question Find the number whose double is \(60\) more than its half.
  \question The internal angles of a heptagon sum to \(900^\circ\). Find
  the sum of the internal angles of a heptadecagon.
  \question Given that \[6^{2x+y}=36^7\] and \[6^{x+4y}=216^7\] find \(xy\).
  \question Two vertices of an equilaterial triangle are \((1,1)\)
  and \((7,1)\). Find the coordinates of the third vertex.
  \question What is the thousandth digit after the decimal
  place of \(1/7\)?
  \question Evaluate \[3+6+9+\cdots+300.\]
  \question How many four-digit numbers can be formed using the digits \(3,1,4,1\)?
  \question A right-angled triangle has vertices \((0,0)\), \((a,b)\),
  \((a,0)\) with hypotenuse \((a+b)/(a-b)\). The point \((a,b)\) lies
  on a circle of radius \(4\). Find \(ab\).
  \question In a set of \(50\) numbers, the average of the first \(20\) is \(30\) while
  the average of the other \(30\) is \(20\). What is the average of all \(50\) numbers?
  \question Find the first date in this millenium to have \(8\) unique
  digits when written in the form DD/MM/YYYY.
  \question Find the probability that a randomly chosen number between
  \(2\) and \(100\) inclusive is prime.
  \question A family consists of a mother, a father, and a number of children. The average
  age of the family is \(20\). The average age of just the mother and the children is \(16\).
  If the father's age is \(48\), how many children are there in the family?
  \question An equilaterial triangle with area \(1\) is inscribed inside a circle.
  This circle is itself inscribed inside a larger equilaterial triangle. What is the area
  of the larger triangle?

  \question What is the smallest positive integer that leaves a remainder of \(1\)
  when divided by each of the numbers \(2,3,\ldots,10\)?
  \question At exactly midnight the hour hand of a clock begins to move twice its normal speed
  while the minute hand begins to move at half its normal speed. What is the correct time the next
  time the two hands meet?
  \question A regular decagon has an area of \(100\) square units.
  What is the length of one of its sides to three decimal places?
  \question If \[A=1^{-3}+2^{-3}+3^{-3}+\cdots\] find \[1^{-3}+3^{-3}+5^{-3}+\cdots\] in terms of \(A\).
  \question Today, the 18th of July, 2023 is a Tuesday. What was the day of the week on the 18th of July 1823?
  \question The numbers \(1,2,\ldots,1000\) are written on the whiteboard. On each turn, two numbers \(x\)
  and \(y\) erased from the board and replaced by \(xy-x-y+2\). This continues until only one number remains. What is this number?
  \question Two real  numbers \(x\) and \(y\), where \(x>y\) have the property that they are equal to \(1\) plus their recriprocal. What is \[\frac{x^2-y^2}{\sqrt5}?\]
\end{questions}
\pagebreak
\begin{enumerate}
\item \(40\)
\item \(2700^\circ\)
\item \(20\)
\item \((4,1+3\sqrt3)\)
\item \(8\)
\item \(15150\)
\item \(12\)
\item \(24/5\)
\item \(24\)
\item 17/06/2345
\item \(25/99\)
\item \(6\)
\item \(4\)
\item \(2521\)
\item \(3\) am
\item \(3.605\)
\item \((1-2^{-3})A\)
\item Friday
\item \(1\)
\item \(1\)
\end{enumerate}
\end{document}