\documentclass[a4paper,12pt,addpoints]{exam}
\usepackage{amsmath}
\usepackage[a4paper,pass]{geometry}
\begin{document}
\section*{Year 11/12}
\pagestyle{empty}
\begin{questions}

\question[3]
There are twelve lockers, numbered from \(110\) to \(121\). The keys
to these twelve lockers are numbered \(1\) to \(12\). 
Each locker number is divisible by the number on its key.

Determine the key number for each locker.

\droptotalpoints

\question
\begin{parts}
    \part[1] Two lines can divide the plane into at most how many regions?
    \begin{solution}
        \(4\)
    \end{solution}
    \part[1] Three lines can divide the plane into at most how many regions?
    \begin{solution}
        \(7\) (If you try it on paper, you will see that the third line can only intersect
        the existing \(2\) lines in at most \(2\) places, resulting in \(3\) new regions.
        Therefore the number of regions is \(4+3=7\).)
    \end{solution}
    \part[1] If we add a fourth line, what is the maximum number of intersection
    points that line can make with the existing lines?
    \begin{solution}
        \(3\)
    \end{solution}
    \part[1] Four lines can divide the plane into at most how many regions?
    \begin{solution}
        \(11\)
    \end{solution}
    \part[3] What is the maximum number of regions that \(n\) lines can divide the plane into?
    \begin{solution}
        \(1+n(n+1)/2\)
    \end{solution}

  \end{parts}
  \droptotalpoints

\question
\begin{parts}
    \part[1] How many ways are there to create a two-element subset from the set \(\{1,2,3,4\}\)?
    \begin{solution}
        There are \(\binom{4}{2}=6\) ways.
    \end{solution}
    \part[2] How many ways are there to split the set \(\{1,2,3,4\}\) into two disjoint nonempty sets? (Disjoint means the two sets share no elements.) For example, if we had the set \(\{1,2,3\}\) a valid
    splitting would be \(\{1,2\}\) and \(\{3\}\).
    \begin{solution}
        The valid ways are: \(\{1\}\) and \([\{2,3,4\}\), \(\{2\}\) and \(\{1,3,4\}\),  \(\{3\}\) and
        \(\{1,2,4\}\), \(\{4\}\) and \(\{1,2,3\}\), \(\{1,2\}\) and \(\{3,4\}\), \(\{1,3\}\) and \(\{2,4\}\), and \(\{1,4\}\) and \(\{2,3\}\). So \(7\) ways.
    \end{solution}
    \part[2] How many ways are there to split the set \(\{1,2,3,4,5\}\) into two disjoint nonempty sets?
    \begin{solution}
        By listing all the ways again, we find that there are \(15\) ways.
    \end{solution}
    \part[2] Denote by \(S(n,k)\) the number of ways to split a set of \(n\) elements into \(k\) disjoint
    nonempty sets. Suppose we know what \(S(n-1,2)\)  is. What happens when we add another element
    to the set? What happens to each of the \(S(n-1,2)\) ways to do the partitioning?
    \begin{solution}
        For each way to do the partitioning, we may add the new number \(n\) to either
        set to obtain a new partitioning. There is also one new partition that is not
        obtained by adding \(n\) to existing one, and that is the splitting into \(\{n\}\)
        and \(\{1,2,\ldots,n-1\}\). Therefore we have \[S(n,2)=2S(n-1,2)+1.\]
    \end{solution}
    \part[2] What is \(S(n,2)\)?
    \begin{solution}
        We have \(S(1,2)=0\), \(S(2,2)=1\), \(S(3,2)=3\), \(S(4,2)=7\), and \(S(5,2)=15\).
        From this pattern and the recursive formula found above it is easy to prove by
        induction that \(S(n,2)=2^{n-1}-1.\)
    \end{solution}
    \bonuspart[3] If there are \(S(n-1,k-1)\) ways to partition a set of \(n-1\) elements into \(k-1\) nonempty
    disjoint sets, what is \(S(n,k)\) in terms of \(S(n-1,k)\) and \(S(n-1,k-1)\)?
    \begin{solution}
        We need to think about what happens when we add the \(n\)th element. Clearly we may
        put this object into a separate set by itself and partition the other \(n-1\) objects
        into \(k-1\) nonempty disjoint subsets, because all up there would be \(k\) sets. 
        There are \(S(n-1,k-1)\) ways to do this. We may also take the \(n-1\) objects
        and partition them into \(k\) sets and add the \(n\)th element to any one of them
        to get another valid partitioning. How many ways are there to do this? In each
        of the \(S(n-1,k)\) ways to partition \(n-1\) objects into \(k\) nonempty sets, there
        are \(k\) different subsets we can put the \(n\)th object into. Therefore
        in this case there are \(kS(n-1,k)\) ways. In total then we have \[S(n,k)=S(n-1,k-1)+kS(n-1,k).\]
    \end{solution}

  \end{parts}
  \droptotalpoints

       \question
\begin{parts}
    \part[2] Is the sum or difference of two rational numbers always rational? Why or why not?
    \begin{solution}
        Yes, because \(a/b+c/d=(ad+bc)/bd\) and both \(ad+bc\) and \(bd\) are integers.
    \end{solution}
    \part[2] Is the product or quotient of two rational numbers always rational? Why or why not?
    \begin{solution}
        Yes, because \((a/b)(c/d)=ac/bd\) and both \(ac\) and \(bd\) are integers.
    \end{solution}
    \part[2] When one rational number is raised to another is the result guaranteed to be rational?
    \begin{solution}
        No; an example is \(2^{1/2}\). 
    \end{solution}
    \part[3] It is known that \(\sqrt2\) is irrational, but it is not known whether \(\sqrt2^{\sqrt2}\) is rational or irrational. 
    Do there exist two irrational numbers such that when one is raised to the other the result is rational? Prove your conjecture.
    \begin{solution}
        The answer is yes.
        Let \(A=\sqrt2^{\sqrt2}\). We don't know if \(A\) is rational or irrational, but it doesn't
        matter. If \(A\) is rational, then it is an example of a rational number that is an
        irrational number raised to another irrational number. If \(A\) is irrational,
        then \(A^{\sqrt2}=\sqrt2^{2}=2\) and this number is the desired example.
    \end{solution}

  \end{parts}
  \droptotalpoints

    \question % OK, this is a monster problem... this'll be the last one on the senior questions.
\begin{parts}
\part[1] If two normal six-sided dice are rolled, what is the probability that the sum of the two numbers is \(2\)?
\begin{solution}
    Out of the \(36\) possible pairs the only one that has a sum of \(2\) is \((1,1)\), so 
    the probability is \(1/36\).
\end{solution}
\part[1] If two normal six-sided dice are rolled, what is the probability that the sum of the two numbers is \(7\)?
\begin{solution}
    There are \(6\) ways to get a \(7\): \((1,6)\), \((2,5\), \((3,4)\), \((4,3)\), \((5,2)\), and
    \((6,1)\). Therefore the probability is \(1/6\).
\end{solution}
\part[2] Consider a biased six-sided die in which the probability of rolling \(n\) is \(p_n\). This die is rigged
so that when two of them are rolled, every possible sum of the two numbers is equally likely. What is \(p_2\) in
terms of \(p_1\)?
\begin{solution}
    The probability of getting a \(3\) is \(p_2p_1+p_1p_2\), which by hypothesis is equal
    to the probability of getting a \(2\), or \(p_1^2\). Therefore \(2p_2=p_1\), which implies
    that \(p_2=p_1/2\).
\end{solution}
\part[1] What is \(p_3\) in terms of \(p_1\)?
\begin{solution}
    Following the previous procedure we find that \(p_3=3p_1/8\).
\end{solution}
\part[2] What is \(p_4\) in terms of \(p_1\)? Do you notice a pattern?
\begin{solution}
    \(p_4=5p_1/16\). The pattern is \(1/2\), \((1/2)(3/4)\), \((1/2)(3/4)(5/6)\), and so on.
\end{solution}
\part[2] Let \(p_{i}=a_{i-1}p_1\). So \(a_0=1\), \(p_2=a_1p_1\), \(p_3=a_2p_1\), \(p_4=a_3p_1\) and so on. What are
\(a_0a_1+a_1a_0\), \(a_0a_2+a_1a_1+a_2a_0\), \(a_0a_3+a_1a_2+a_2a_1+a_3a_0\), etc. all equal to?
\begin{solution}
    If \(p_1p_2+p_2p_1=p_1^2\) then \[a_0p_1a_1p_1+a_0p_1a_1p_1=p_1^2\] which implies 
    \[(a_0a_1+a_1a_0)p_1^2=p_1^2.\] Therefore \(a_0a_1+a_1a_0=1.\) Similarly the other 
    combinations of \(a\)'s can be found to be \(1\) too.
\end{solution}
\part[1] Since we have a six-sided die, \[p_1+p_2+\cdots+p_6=1.\] What is \(p_1\) in terms of \(a_0,a_1,a_2,\ldots,a_5\)?
\begin{solution}
    Since the die is six-sided, we have \(p_1+p_2+\cdots+p_6=1\). We can express this 
    in terms of \(p_1\): \[a_0p_1+a_1p_1+a_2p_1+\cdots+a_5p_1=1\] and hence
    \[ p_1=\frac{1}{a_0+a_1+\cdots+a_5}.\]
\end{solution}
\part[2] Let \(s_i=a_0+a_1+\cdots+a_i.\) What are \(s_0,s_1,s_2\)? What is \(s_5\)?
\begin{solution}
    \(s_0=a_0=1\), \(s_1=a_0+a_1=3/2\), \(s_2=1+1/2+3/8=15/8\). Note that 
    \(s_2=(3/2)(5/4)\), \(s_3=(3/2)(5/4)(7/6)\), \(s_4=(3/2)(5/4)(7/6)(9/8)\) and so on.
    Therefore \[s_5=\frac{3\cdot5\cdot7\cdot9\cdot11}{2\cdot4\cdot6\cdot8\cdot10}=\frac{693}{256}.\]
\end{solution}
\part[3] Is such a six-sided die possible?
\begin{solution}
    From the previous part we find that \(p_1=256/693.\) This means that the probability of
    rolling two of these dice and getting a sum of \(2,3,\ldots,12\) must all be
    \(p_1^2=65536/480249.\) But this is impossible because there are \(11\) possible
    sums and if they are equally likely each must have a probability of \(1/11\). Therefore
    such a die is not possible.
  \end{solution}

\end{parts}
\droptotalpoints

\question
\begin{parts}
    \part[1] What is \(1+2+3+\cdots+200\)?
    \begin{solution}
        \(10100\)
    \end{solution}
    \part[2] Find a formula for \(1+2+\cdots+n\) and prove it.
    \begin{solution}
        The formula is \[1+2+\cdots+n=\frac{n(n+1)}{2}.\] It is easily proved by induction.
        When \(n=1\) we have \(1=1(1+1)/2\) so the base case is done. Now assume that
        \[1+2+\cdots+k=\frac{k(k+1)}{2}\] for some natural number \(k\). Add \(k+1\) to
        both sides: \[1+2+\cdots+k+(k+1)=\frac{k(k+1)}{2}+\frac{2(k+1)}{2}=\frac{(k+1)(k+2)}{2}\] 
        and the induction is complete.
    \end{solution}
    \part[1] Define \[F_k(n)=1^k+2^k+\cdots+n^k.\] (So in the previous part you found a formula for
    \(F_1(n)\).) More compactly it may be written as
    \[F_k(n)=\sum_{i=1}^{n}i^k.\]
    What is \[1^3+(2^3-1^3)+(3^3-2^3)+\cdots+(100^3-99^3)?\]
    \begin{solution}
        This sum is an example of a telescoping sum. All the terms vanish except for
        the \(100^3\) term in the last set of parentheses, so the answer is \(100^3\).
    \end{solution}
    \part[2] By considering \[1^3+\sum_{i=1}^{n}((i+1)^3-i^3)\] find a formula for \(F_2(n)\).
    \begin{solution}
    We know that \[1^3+\sum_{i=1}^{n}((i+1)^3-i^3)=(n+1)^3.\] Expanding both sides
    using the Binomial Theorem we get \[1^3+\sum_{i=1}^{n}(3i^2+3i+1)=n^3+3n^2+3n+1.\]
    After removing the \(1\) from both sides, on the LHS we can break up the sum and pull out the constant factors, like so:
    \[3\sum_{i=1}^ni^2+3\sum_{i=1}^ni+\sum_{i=1}^n1=n^3+3n^2+3n.\]
    But note that \(\sum_{i=1}^ni^2=F_2(n)\) and \(\sum_{i=1}^ni=F_1(n)\), so we have
    \[3F_2(n)+3F_1(n)+n=n^3+3n^2+3n.\] Then we just need to rearrange for \(F_2(n)\) after
    substituting in \(F_1(n)=n(n+1)/2\) to
    find that \[F_2(n)=\frac{2n^3+3n^2+n}{6}.\]
    \end{solution}
    \part[2] Find a formula for \(F_3(n)\).
    \begin{solution}
        Using the same method as above, we find that \[F_3(n)=\frac{n^4+2n^3+n^2}{4}.\]
    \end{solution}
    \part[1] The Binomial Theorem states that \[(x+y)^n=\sum_{k=0}^{n}\binom{n}{k}x^ky^{n-k}.\] If 
    \[(i+1)^k-i^k=\sum_{j=0}^{x}\binom{k}{j}i^j\] what is \(x\) in terms of \(k\)?
    \begin{solution}
        The leading term vanishes, so instead of going from \(0\) to \(k\) the loop only
        goes up to \(k-1\). Therefore \(x=k-1.\)
    \end{solution}
    \part[3] By considering the general telescoping sum 
    \[1^k+\sum_{i=1}^n((i+1)^k-i^k)=(n+1)^k\] show that 
    \[F_{k-1}(n)=\frac{(n+1)^k-1}{k}+\frac{1}{k}\sum_{i=0}^{k-2}\binom{k}{i}F_i(n).\]
    \begin{solution}
        Expand the \((n+1)^k\)  term using the Binomial Theorem, applying the previous part:
        \[(n+1)^i=1^k+\sum_{i=1}^{k}\sum_{j=0}^{k-1}\binom{k}{j}i^j.\]
        Now expand the inner sigma notation:
        \[(n+1)^k=1^k+\sum_{i=1}^{k}\left(\binom{k}{0}i^0+\binom{k}{1}i^1+\binom{k}{2}i^2+\cdots+\binom{k}{k-1}i^{k-1}\right).\]
        We may take out the binomial coefficients, as they are independent of the sum variable
        \(i\), so we have 
        \[(n+1)^k=1^k+\binom{i}{0}\sum_{i=1}^ki^0+\binom{k}{1}\sum_{i=1}^ki^1+\binom{k}{2}\sum_{i=1}^ki^2+\cdots+\binom{k}{k-1}\sum_{i=1}^{k}i^{k-1}.\]
        We're interested in \(F_{k-1}(n)\), so we notice that the binomial coefficient
        \(\binom{k}{k-1}=k\) and isolate that term. The other sums are \(F_0(n)\), \(F_1(n)\),
        \(F_2(n)\) and so on. Therefore we get
        \[(n+1)^k=1^k+\binom{k}{0}F_0(n)+\binom{k}{1}F_1(n)+\binom{k}{2}F_2(n)+\cdots+\binom{k}{k-2}F_{k-2}(n)+kF_{k-1}(n)\]
        or more compactly,
        \[(n+1)^k=1+\sum_{i=0}^{k-2}\binom{k}{i}F_i(n)+kF_{k-1}(n).\]
        Rearranging for \(F_{k-1}(n)\) we get the desired result.
    \end{solution}

  \end{parts}
  \droptotalpoints

\end{questions}
Total: \numpoints
\end{document}
