\documentclass[a4paper,12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[bf,tiny]{titlesec}
\usepackage{fancyhdr}

\usepackage[indent=0pt,skip=10pt]{parskip}

\usepackage{amsmath}
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\usepackage{amssymb}
\let\mathbbalt\mathbb

\usepackage{fontspec}
\usepackage{unicode-math}
\let\mathbb\mathbbalt

\newcommand{\naturals}{\mathbb{N}}
\newcommand{\reals}{\mathbb{R}}
\newcommand{\rationals}{\mathbb{Q}}
\newcommand{\integers}{\mathbb{Z}}

\newcounter{problemno}
\setcounter{problemno}{1}
\newcommand{\problem}{\section*{Problem \theproblemno}\refstepcounter{problemno}}
\newcounter{questionno}
\setcounter{questionno}{1}
\newcommand{\question}{\par\noindent\thequestionno.\enspace\refstepcounter{questionno}}

\newcommand{\thetitle}{First Meeting Problems}

\pagestyle{empty}
\begin{document}
Melbourne High School\\
Maths Extension Group 2023\\
\textbf{\thetitle}\\

\question
Solve the equation \(4t/3+2=2(t-2)\).
\question
A square circumscribes a circle of radius \(a\). Find the probability that a randomly chosen
point in the square is also in the circle.
\question
Find a number that is the same when added to \(5\) as when multiplied by \(5\).
\question
Evaluate
\[
\frac{1}{1\cdot2}+\frac{1}{2\cdot3}+\frac{1}{3\cdot4}+\frac{1}{4\cdot5}+\cdots+\frac{1}{99\cdot100}.
\]
\question
Call an integer to be $gobbus$ if every digit is either \(0\) or \(2\) and it is divisible by \(15\). What is the smallest number that is gobbus?
\question
I have three jars of cookies. The product of the number of cookies in each jar is \(36\). Unfortunately, even
if I tell you the sum of the number of cookies in each jar, you won't be able to determine how many there are
in each jar, but I do know that the jar with the most cookies is currently open. How many cookies are there
in each jar?
\question
If \(1/(a+b)=4\) and \(a^2+b^2=16\) find \(1/a+1/b.\)
\question
Find all primes less than \(100\) that are the sum of two squares.
\question
Find all solutions in positive integers \(x\) and \(y\) of the equation
\(15x^2+2y^2=100+13xy.\)
\question
 Find the last digit of $2^{2023}$.
\question
If $a,b,c > 0$ and $a+b+c=1$, prove that $a^2+b^2+c^2+1\ge4(ab+bc+ca)$.
\question
Define the $alternating$ $sum$ of a set of positive integers as follows: List the elements of the set in decreasing order. Take the first number in the list, subtract the second, add the third, subtract the fourth, and so on. For example, the alternating sum of the set ${3,5,11,7}$ is $11-7+5-3=6$. Find the sum of the alternating sums of all the subsets of $\{1,2,3,4,5,6,7,8,9,10\}$.
\question
Evaluate \(5^{261}\pmod{397}\).
\question
Suppose $BP$ and $CQ$ are bisectors of the angles $B$, $C$ of triangle ABC and suppose $AH$, $AK$ are the perpendiculars from $A$ to $BP$, $CQ$. Prove that $KH$ is parallel to $BC$.
\question
The Fibonacci sequence is defined by the recurrence relation \(F_n=F_{n-1}+F_{n-2}\), where \(F_0=0\) and \(F_1=1\).
Consider the power series
\[
\begin{split}
G(x)&=\sum_{i=0}^\infty F_ix^i\\
&=F_0x^0+F_1x^1+F_2x^2+F_3x^3+F_4x^4+F_5x^5+F_6x^6+\cdots.
\end{split}
\]
Find a closed-form expression for \(G(x)\).
\end{document}
